---
title: "Uncertainty"
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

### Introduction

Estimates have uncertainty. Here we look at the mean and its sampling distribution for a continuous and a binary scaled variables. We calculate the mean and its standard error using the normal approximation and using the bootstrap.  

### Installing the packages and loading the data

To illustrate the technique, we are going to use data from eBeer. 

```{r message=FALSE, warning=FALSE}
# install.packages(c("memisc", "foreign", "data.table", "dplyr","VGAM"))
# options("scipen"=100, "digits"=4)

# install.packages(c("memisc", "foreign", "data.table", "dplyr","VGAM"))

library(data.table)
library(foreign)
library(memisc)
```

Set the working directory to a folder where you will you put class related stuff. 

```{r} 
setwd("C:/Users/gknox/Dropbox/Customer Analytics/R notebooks/RFM & logistic")
```

Set the seed so that random draws are the same for everyone.  I create a folder within the working directory for data called ... data.  

```{r}
set.seed(19103)

ebeer <- data.frame(as.data.set(spss.system.file('./data/E-Beer.sav')))

#ebeer<- data.frame(as.data.set(spss.system.file(file.choose())))
```

Let's take a look at the variable monetary value.  This is the average amount per transaction spent by customer.  I'm going to refer to it as the monetary value.  Here's its histogram.

```{r}
dim(ebeer)

head(ebeer)

par(mai=c(.9,.8,.2,.2))  
hist(ebeer$M, main="Monetary", xlab="Monetary amount")
```

How many customers are in our data?  What is the mean monetary value, its variance, and standard error?  

Remember the sample mean is:
$$
\bar{x} = \frac{1}{n} \sum_{i=1}^{n} x_i
$$
And the variance of it is:
$$
var(\bar{x}) = var \left( \frac{1}{n} \sum_{i=1}^{n} x_i \right) =  \frac{1}{n^2} \sum_{i=1}^{n} var(x_i) =  \frac{\sigma^2}{n}
$$
where $\sigma^2 = \frac{1}{n-1}\sum_{i=1}^{n} (x_i - \bar{x})^2$ is the variance. The variance goes inside the parentheses because we assume each customer is independent. And the standard error is $se(\bar{x}) = \frac{\sigma}{\sqrt{n}}$. (I'm ignoring finite sample adjustment here.)

```{r}
nrow(ebeer)
xbar <- mean(ebeer$M)
xbv <- var(ebeer$M)/nrow(ebeer)
xbse <-  sqrt(xbv)

xbar
xbv
xbse
```

What is the sampling distribution of the mean?  If we were able to get a new sample from the same data-generating process what would it be?  According to the central limit theorem, the distribution is normal, $\bar{x} \sim N(\bar{x}, var(\bar{x}))$. This distribution below:

```{r}

xx <- seq(89,95,length=1000)

par(mai=c(.9,.8,.2,.2))
plot(xx, dnorm(xx, xbar, xbse), type="l", col="royalblue", lwd=1.5,
     xlab="average monetary value", ylab="density")
```

Another way to compute the distribution is using the bootstrap.  The idea is to resample your data *with replacement*, calculating the statistic each time, e.g., $\bar{x}_1, \bar{x}_2, \dots \bar{x}_B$ and using that to calculate the distribution of the statistic.  

```{r}
# nonparametric bootstrap
B <- 10000
mub <- c()
for (b in 1:B){
  samp_b = sample.int(nrow(ebeer), replace=TRUE)
  mub <- c(mub, mean(ebeer$M[samp_b]))
}
sd(mub)

par(mai=c(.8,.8,.2,.2))
hist(mub, main="", xlab="average monetary value", breaks = 100,
     col=8, border="grey90", freq=FALSE)
lines(xx, dnorm(xx, xbar, xbse), col="royalblue", lwd=1.5)
```

Compare the standard errors from the normal approximation and the bootstrap sample.  You can create a 95% confidence interval around the mean using the standard error from the bootstrap.

```{r}
mean(mub) + c(-1,1)*2*sd(mub)
```

The bootstrapped sampled with replacement.  In the last sample, you can see person 1 was sampled twice, person 2 also twice, etc.:

```{r}
sort(samp_b)[1:10]
```


### Help for assignment 1

Now let's take a look at a binary variable, marketing response. 

```{r}
table(ebeer$respmail)

mean(ebeer$respmail)
```

You may get a problem here because only about half of the people in the data set received a mail. For those missing a mail, the response will be NA, which creates an error. So let's remove those people, focusing on the subset of people who received a mailing.

```{r}
ebeer<-ebeer[!is.na(ebeer$respmail),]

n<-nrow(ebeer)
```

Make sure response is numeric, 1 if responded and 0 if not, and not a factor variable with two labels.  The function as.numeric gets used a lot.

```{r}
ebeer$respmail<-as.numeric(ebeer$respmail)

table(ebeer$respmail)
```

In the case that x is binary, the variance is $\sigma^2 = \hat{p}(1-\hat{p})$. 

